Polynomial Time and Dependent Types

### 2023-10-24 paper/paper.tex
[ paper ] PP/BPP and semantic mistakes

For this section, the affine linear $\lambda$-calculus we will use
will have linear functions and $\otimes$-products:\bob{CHECK THAT THIS CHANGE TO THE VAR RULE IS OK, MENTION THAT EXCHANGE IS IMPLICIT}
\begin{mathpar}
  \inferrule*
  { }
  {\Gamma, x : A \vdash x : A}

@@

\subsubsection{Probabilistic Polynomial Time}

\bob{Check PP vs BPP. Also cite Dal Lago PP-time paper}

By changing the computation effects supplied to a program, we can
change the complexity class. To capture Probabilistic Polynomial time,
we use a (non-iterable) data structure representing trees of
probabilistic choices, where $\mathbb{Q}[0,1]$ is some type of
(non-iterable) rationals in the closed interval $[0,1]$:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{Dist}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{Dist}\,A \\
    \quad \mathsf{choice} : \mathbb{Q}[0,1] \to (\BoolTy\ to \mathrm{Dist}\,A) \to \mathrm{Dist}\,A
  \end{array}
\end{displaymath}
As in the non-deterministic case, a function $A \to \mathrm{Dist}\,B$
in the $\sigma = 1$ fragment is a polytime probabilistic
computation. Again, the use of a function type here ensures that each
branch of the tree is constructable in polynomial time, not the whole
tree. In the $\sigma = 0$ fragment we can write a function that
computes the probability of a $\mathrm{Dist}\,\BoolTy$ computation
being true:
\begin{displaymath}
  \mathrm{probTrue} \stackrel0: \mathrm{Dist}\,\BoolTy \to \mathbb{Q}[0,1]
\end{displaymath}
We can now define the class of probabilistic polynomial time decision
problems, where the decider is allowed to make probabilistic choices
as long as it is correct with probability at least $\frac{1}{3}$:
\begin{displaymath}
  \mathrm{PP}(A,P) =
  \begin{array}[t]{l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{Dist}(\BoolTy))) \, \otimes
    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{1}{3}) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Again, problems in PP are easily seen to be closed under polytime
reductions.

@@

\citet{atkey18qtt} described a general class of QTT models termed
\emph{Quantitative Categories with Families} (QCwFs). This class of
models fits well with the syntax of QTT, but to actually build models
it is convenient to construct them from certain indexed linear
preorders. The BCI-realisability model construction given by
\citet{atkey18qtt} is a special case of this construction, so we only
sketch a special case here to concentrate on the specifics of
complexity bounded realisability. \bob{state clearly that Atkey 2018
  has a mistake, and this construction here is more general}

\begin{definition}\label{def:indexed-linear-preorder}
  A \emph{$\mathbb{N}$-linear preorder} is a preordered set
  $(L, \leq)$ with
  \begin{enumerate}
  \item a commutative monoid $(I, -\otimes-)$ that is monotone
    w.r.t. the order;
  \item is closed: there is an operation
    $\multimap : L \times L \to L$ such that $x \otimes y \leq z$ iff
    $x \leq y \multimap z$; and
  \item has a function $\oc : \mathbb{N} \to L \to L$, to interpret
    resource requirement adjustments, satisfying:
    \begin{enumerate}
    \item $\oc_0 X \simeq I$, for discarding;
    \item $\oc_{m+n} X \leq (\oc_m X) \otimes (\oc_n X)$, for duplication;
    \item $\oc_m\oc_n X \leq \oc_{mn} X$ for nesting;
    \item $\oc_1 X \leq X$ for extraction / dereliction;
    \item $(\oc_n X) \otimes (\oc_n Y) \leq \oc_n(X \otimes Y)$, for distribution; and
    \item $n \leq m$ implies $\oc_n X \leq \oc_m X$, for usage weakening.
    \end{enumerate}
  \end{enumerate}
  The collection of all linear preorders and functions that preserve the
  order and the operations forms a category $\LinPreorder$.

@@

$\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
action of $(\mathbb{N}, \leq)$ on $M$ defined above. The modality
$\oc_n$ has no effect on realising values. It only serves to alter the
resource potentials.\bob{fix this to match the Agda implementation: special case $0$}
\begin{displaymath}
  \oc_n\,X = \{(a,\alpha,v) \mid \exists \alpha'.\,M(n \cdot \alpha', \alpha) = 0 \land (a, \alpha', v) \in X \}
\end{displaymath}

>>>>

For this section, the affine linear $\lambda$-calculus we will use
will have linear functions and $\otimes$-products. Contexts are
treated up to permutation, so uses of exchange are implicit.
\begin{mathpar}
  \inferrule*
  { }
  {\Gamma, x : A \vdash x : A}

@@

\subsubsection{Bounded-error Probabilistic Polynomial Time}

By changing the computation effects supplied to a program, we can
change the complexity class. To capture the class BPP of Bounded-error
Probabilistic Polynomial time \cite{aurora}, we use a (non-iterable)
data structure representing trees of probabilistic choices, where
$\mathbb{Q}[0,1]$ is some type of (non-iterable) rationals in the
closed interval $[0,1]$:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{Dist}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{Dist}\,A \\
    \quad \mathsf{choice} : \mathbb{Q}[0,1] \to (\BoolTy\ to \mathrm{Dist}\,A) \to \mathrm{Dist}\,A
  \end{array}
\end{displaymath}
As in the non-deterministic case, a function $A \to \mathrm{Dist}\,B$
in the $\sigma = 1$ fragment is a polytime probabilistic
computation. Again, the use of a function type here ensures that each
branch of the tree is constructable in polynomial time, not the whole
tree. In the $\sigma = 0$ fragment we can write a function that
computes the probability of a $\mathrm{Dist}\,\BoolTy$ computation
being true:
\begin{displaymath}
  \mathrm{probTrue} \stackrel0: \mathrm{Dist}\,\BoolTy \to \mathbb{Q}[0,1]
\end{displaymath}
We can now define the class of probabilistic polynomial time decision
problems, where the decider is allowed to make probabilistic choices
as long as it is correct with probability at least $\frac{2}{3}$:
\begin{displaymath}
  \mathrm{BPP}(A,P) =
  \begin{array}[t]{l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{Dist}(\BoolTy))) \, \otimes
    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{2}{3}) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Again, problems in BPP are easily seen to be closed under polytime
reductions.

Probabilistic Polynomial time has previously been considered in the
setting of implicit computational complexity by
\citet{dallago_et_al:LIPIcs.MFCS.2021.35} and \citet{LagoT15}. In both
cases, they must build probabilistic choice into the language, and
have difficulty in directly capturing the class BPP due to its
semantic nature \cite{aurora}, where the correctness of implementation
is probabilistic. With a dependently-typed host language, adding
probabilistic choice as an effect, and capturing the semantic
constraint of BPP is straightforward.

@@

\citet{atkey18qtt} described a general class of QTT models termed
\emph{Quantitative Categories with Families}
(QCwFs). \citet{atkey18qtt} constructs QCwFs from certain Linear
Combinatory Algebras (LCA), where terms in the $\sigma = 1$ fragment
are realised by elements of the LCA. However, there is a mistake in
that paper where the interpretation of contexts is stated to be the
category of \emph{assemblies} over the LCA, where it ought to be the
category of sets paired with realisability relations, with no
guarantee that all elements be realisable.

Here, we fix the mistake of \citet{atkey18qtt}, and provide a more
general construction of QCwFs in terms of indexed linear preorders. We
construct indexed linear preorders specific to our polytime setting
below. They could also be constructed from LCAs.

\begin{definition}\label{def:indexed-linear-preorder}
  A \emph{$\mathbb{N}$-linear preorder}\footnote{We specialise to the semiring $\mathbb{N}$ here, but the same definition works for any suitable semiring $\mathcal{R}$.} is a preordered set
  $(L, \leq)$ with
  \begin{enumerate}
  \item a commutative monoid $(I, -\otimes-)$ that is monotone
    w.r.t. the order;
  \item is closed: there is an operation
    $\multimap : L \times L \to L$ such that $x \otimes y \leq z$ iff
    $x \leq y \multimap z$; and
  \item has a function $\oc : \mathbb{N} \to L \to L$, to interpret
    resource requirement adjustments, satisfying:
    \begin{enumerate}
    \item $\oc_0 X \simeq I$, for discarding;
    \item $\oc_{m+n} X \leq (\oc_m X) \otimes (\oc_n X)$, for duplication;
    \item $\oc_m\oc_n X \leq \oc_{mn} X$ for nesting;
    \item $\oc_1 X \leq X$ for extraction / dereliction;
    \item $(\oc_n X) \otimes (\oc_n Y) \leq \oc_n(X \otimes Y)$, for distribution; and
    \item $n \leq m$ implies $\oc_n X \leq \oc_m X$, for usage weakening.
    \end{enumerate}
  \end{enumerate}
  The collection of all linear preorders and functions that preserve the
  order and the operations forms a category $\LinPreorder$.

@@

$\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
action of $(\mathbb{N}, \leq)$ on $M$ defined above. When $n > 0$, the
modality $\oc_n$ has no effect on realising values. It only serves to
alter the resource potentials. In the $n = 0$, case the realising
value must be $*$, in order to satisfy the $\oc_0 X \cong I$ condition
in \defref{def:indexed-linear-preorder} 3(a):
\begin{displaymath}
  \begin{array}{lcl}
    \oc_0\,X&=&\{(a, \alpha, *) \mid a \in A, \alpha \in M \} \\
    \oc_n\,X&=&\{(a,\alpha,v) \mid \exists \alpha'.\,M(n \cdot \alpha', \alpha) = 0 \land (a, \alpha', v) \in X \}
  \end{array}
\end{displaymath}
